### 2019-06-14 Manuscripts/cog-psych/post-revision/post-cogpsych.tex
Massive update with Corner condition writeup. Still rough but probably human readable.

Creating and Learning Something Different: \\
Similarity, Contrast, and Representativeness in Categorization

@@

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most fascinating
aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking the extent that the contrast
between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization. Our work also serves as a concrete example of
how well-established categorization models can be applied beyond category learning.

@@

% % OLD OLD INTRO:
% Creativity, innovation, imagination, and the creation of new ideas are among the most fascinating, important and difficult human capabilities to study. They are fascinating and important capabilities because they have produced some of the greatest scientific breakthroughs that revolutionized the world. Unfortunately, technological and scientific breakthroughs are rare and the result of a great deal of cognitive effort, which makes it difficult to investigate the underlying cognitive representations and processes responsible for them using standard experimental methodology or to formalize them in computational models. Although we tend to focus on the most salient products of these processes (e.g., scientific breakthroughs), every person is likely to have generated many novel sentences, thoughts, and/or drawings given the combinatorial explosion of possible statements given a reasonably-sized set of primitive elements and combination rules in language, thought, and perception \citep{goldstone2003}. This observation provides a fantastic opportunity: By examining how people create new concepts in a domain amenable to formalizing in computational models, we can investigate (some aspects of) creativity, innovation, and imagination scientifically.

% Towards the aim of providing a formal account of the cognitive mechanisms involved in creativity and innovation, we explore a form of creative cognition that can be studied using standard behavioral methodology and is amenable to formal modeling: category generation. In a category generation task, participants are given some background knowledge about a domain (a cover story and/or learning one or more categories in the domain) and then are asked to generate another category in the domain. Unlike creativity and innovation, categorization is well studied from a breadth of perspectives, including behavioral, neural, comparative, and computational  \citep{kurtz2015human,mack2013,margolis2015,pothoswills2011}. By analyzing a creative task using formal tools from the categorization literature, we can formalize some theories of creativity and test them empirically in a more rigorous manner.

%cut from submission intro:

%Yet, it is also one of the most difficult human capabilities to study.  We tend to focus on the most salient products of conceptual generation (e.g., scientific breakthroughs), but every person is likely to have generated many novel concepts in ordinary life. Indeed, the generation of novel sentences, works of art, and so on may also be considered products of our capacity to create new concepts. However, while this capacity is of great interest, it is also highly complex, making it difficult to study in a typical laboratory setting. By consequence, we do not have a strong understanding of the representations and processes involved in the generation of new concepts.

%Questions concerning the generation of novel concepts are typically addressed within research on creativity, a tremendously diverse field employing a multitude of experimental and theoretical approaches \citep[see][]{kaufman2010handbook}. The processes underlying conceptual generation are commonly investigated using the creative cognition approach \citep{finke1992creative, smith1995creative}, in which the products of creative acts (such as drawings) are analyzed to obtain insights into how they might have been created. These approaches have made great advances in our understanding of creativity, but the focus on particular types of creative products such as drawings makes it difficult to formalize the proposed processes and representations within a computational modeling framework. As a consequence, this approach does not often employ computational models to provide formal tests of competing theories.

% ------- BEGIN! ------- %
\begin{flushleft}

@@

Creating and learning new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. In fact, computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate and learn new innovations are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create and learn
new categories that are as interesting as cell phones and smart phones? As a
step towards these goals, in this article we investigate the expectations that
people have when they learn and generate new categories using computational
models and behavioral experiments.

@@

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another. To foreshadow, we find
that this notion of contrast predicts human category learning performance, above
and beyond other category learning models in a novel behavioral experiment. This
demonstrates that contrast plays a central role in other important
categorization tasks, and it is not specific to category generation.

@@

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the performance of each model in
capturing human category generation. Afterwards, we find support for category
contrast predicting the ease at which people learn a pair of categories in a
novel empirical experiment. We conclude with a discussion of the implications of
our results for categorization and directions for future
work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and thus more distinct from and less similar to other categories. We
adopt the formalization given by \cite{tenenbaum01}, where the
representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is the
relative amount of evidence that is provided by $y$ for a given hypothesis $h$
in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in H'}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. Generally,
this mechanism of representativeness results in the generation of exemplars that
are less similar to exemplars from other categories. We see this illustrated in
Figure \ref{fig:representative-examples}, where the representativeness model
displays a strong preference for generating exemplars from target category 'B'
that are further away from the contrast category 'A'.

@@

In the sections below, we present two experiments demonstrating systematic
effects of category contrast on category generation inspired by factors
influencing how the contrast models generate new categories. Our experiments are
based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward
translation of the traditional artificial classification paradigm to the task of
category generation: Participants are first exposed to a single,
experimenter-defined category, and are then asked to generate members of a new
category. We then report quantitative analyses of our results comparing the exemplar dissimilarity
and the representativeness accounts of our results to that of the hierarchical
Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}.
After analyzing the category generation studies, we present third experiment to
demonstrate the effect of contrast in category learning. This ensures that our
results are not idiosyncratic to category generation.

@@

%Experiment 3 starts here
\section{Experiment 3}

Experiments 1 and 2 clearly establish the importance of contrast in category
generation. Follow-up model-based analyses illustrated that both contrast models
account for participant performance better than models that do not take into
account contrast. We also found more support for contrast as representativeness
hypothesis over contrast as exemplar dissimilarity hypothesis.

Although category generation is an intriguing aspect of human cognition, it has
played a relatively limited role in the broader categorization literature. To
establish that contrast plays a broader role in categorization, we conducted a
traditional category learning study. In it, participants learned two categories,
where the categories learned by each participant in Experiment 3 was yoked to
the `Alpha' shown and `Beta' generated by each participant in Experiment 2. Our goal
is to investigate whether model performance (in generating categories) is
consistent with human performance (in learning those categories). Specifically, if contrast plays a broader role in categorization, then `Beta' categories rated as "better" according
to the contrast models should be easier for participants to learn.
%\footnote{It is possible to adjust the present models of category generation such that they can make category learning predictions. However, such an adjustment makes the strong assumption that the very same processes governing category generation also apply to category learning. Taking this approach would fall outside the scope of the current investigation.} 
\subsection{Participants, Materials, and Procedure}

Experiment 2 recruited 122 participants who each generated one set of four `Alpha'
and four `Beta' category exemplars. Experiment 3 recruited the same number of
participants.

Participants observed four blocks of eight trials. Each trial began with the
presentation of a fixation cross for 500 ms. This was followed by the
presentation of one exemplar randomly sampled without replacement from the
unique category set. Participants were tasked with assigning the presented
exemplar to either the `Alpha' or `Beta' category with no time limit imposed.
Feedback was automatically displayed for 2500 ms after each response.

\subsection{Results and Analysis}

Among the 122 originally generated category sets were 102 unique sets (i.e.,
they contained a unique collection of `Alpha' and `Beta' exemplars). Consequently,
for our subsequent analyses for this experiment we only use data from the first
102 participants that were presented with a unique category set.

Overall accuracy of the participants was high, with a mean error rate of $.19$
($SD = .19$). Aggregated error rates for each block are presented in Figure
\ref{fig:learningcurve}.

\begin{figure}
    \begin{center}
    \includegraphics[width=\textwidth/2]{figs/e3-learningcurve.pdf}
    \caption{Average error rate for each successive block. Green discontinuous
line represents the overall mean error rate.}
    \label{fig:learningcurve}
    \end{center}
\end{figure}

In order to analyze the consistency between model and human performance, we
optimized each model such that the Pearson correlation between the model's
negative log-likelihood of generating each unique category set and the
participant's error in learning that category set is maximized. As with the
analysis in Section \ref{section:individual-diff}, we simulated the weighting of
each dimension independently depending on the generated categories.

The greatest correlations were observed when fitting the contrast models.
Specifically, PACKER correlated with human performance with
$r = .65, 95\%\,CI\, [.48,.78]$, and the representativeness
model correlated with human performance with $r = .59, 95\%\,CI\,[.40, .73]$\footnote{Confidence intervals were obtained using a bootstrap
  method with 1000 bootstrap samples.}.
Copy-and-tweak yielded a correlation of $r = .45, 95\%\,CI\,[.28, .62]$ and the
hierarchical Bayesian model yielded a similar correlation of
$r = .46, 95\%\,CI\, [.30, .62]$. These results are graphically presented in Figure
\ref{fig:perror_corr}.

\begin{figure}
    \begin{center}
    \includegraphics[width=\textwidth]{figs/modelvspptp.pdf}
    \caption{Scatterplots indicating the correlations between observed
      participant error and model fit. Blue discontinuous line are best-fit
      lines. Green discontinuous lines indicate boundaries generated by the 95\%
      CI around the fitted correlations. }
    \label{fig:perror_corr}
    \end{center}
\end{figure}

To emphasize the strong influence of contrast in maximizing the association
between category generation and category learning, we computed the correlations
over wide range of $\theta_{contrast}$ values, with the other parameter values
of PACKER held constant at their optimized levels. As presented in Figure
\ref{fig:packer-corr}, the correlation quickly increases with increasing weight
on contrast, reaching a plateau above values of around 4.0.

\begin{figure}
    \begin{center}
    \includegraphics[width=0.8\textwidth]{figs/packer-corr.pdf}
    \caption{Correlation between PACKER's fit and participant error as a
function of the $\theta_{contrast}$ parameter. To facilitate comparison,
PACKER's other parameters ($c$, $\theta_{target}$) were set to the best fitting
values obtained for copy-and-tweak in Table \ref{table:global-model-fits}. Grey
dashed line represents the correlation between Copy \& Tweak's fit and
participant error.}
    \label{fig:packer-corr}
    \end{center}
\end{figure}

@@

We conducted simulations comparing the contrast models' account of our results
to the classical proposals for category generation: a ``copy-and-tweak'' model
(realized as a variant of PACKER with no sensitivity to category contrast), and
a hierarchical Bayesian model designed to explain the classic distributional
similarity effect. In all simulations, we found that the contrast models
captured a previously unexplained and unexplored aspect of human category
generation. In particular, by measuring PACKER's fit as a function of its
prioritization of within- and between-category similarity, we observed that
considering either constraint exclusively results in a relatively low-quality
account. Instead, PACKER's best results were obtained when both constraints are
considered, indicating that human learners do not generate novel concepts
exclusively on the basis of within-category similarity or between
class-contrast. This finding mirrors our behavioral results and demonstrates
that both constraints influence generation when explained with an exemplar
model. Ultimately, our category generation data were best explained by the
representativeness model. The exact nature of its superiority over the other
models is still not yet clear. However, as we explore in a later section, it is
likely that the representativeness mechanism itself (i.e., independent of the
hierarchical Bayesian foundation) captures a core aspect of both category
generation and category learning.

@@

Although the focus of this article has been to address the role of contrast in category generation and learning, our findings and approach have relevant implications
for research in creative cognition. A central focus of the creative cognition
approach has been to explain acts of creativity in terms of the mental
representations and processes that are commonly studied in cognitive psychology
and cognitive science \citep{finke1992creative,smith1995creative}. However,
unlike other fields in the study of cognition, creative cognition research
rarely employs quantitative models to evaluate the explanatory value of such
representations and processes. Our modeling results provide a concrete example
of how formal approaches may be used to gain insight into the nature of creative
cognition.

@@

\subsection{Exemplar Dissimilarity and Representativeness in Categorization} 
Although both PACKER and the representativeness model perform better than the
classical models, they appear to be capturing fundamentally different aspects of
contrast in categorization. Specifically, while the representativeness model
excelled in predicting our category generation data, PACKER was the best-fitting
model in accounting for participant error in category learning. Given the
history of their development, this is perhaps not entirely surprising. PACKER
is implemented as an extension of the GCM, which was originally developed for (and
has been highly successful in) category learning \citep{nosofsky1986attention,
  nosofsky1994comparing}. Conversely, the representativeness mechanism was
conceived as a heuristic describing the relative extent to which an individual
observation is generated from some unobserved population
\citep{kahneman1972subjective,abbott16}. This generative nature of the
representativeness mechanism intuitively lends itself to the prediction of
category generation data.

@@

Ultimately, the source of the discrepancy between the performance of the two
contrast mechanisms in category generation versus category learning is still not
entirely clear. Our results could be an indication of a fundamental difference
between the processes involved where exemplar dissimilarity is a larger factor
in category generation while representativeness exerts more influence in
category learning. Alternatively, it is possible they are both approximating
some factor that we still have not fully identified. While a thorough
investigation is beyond the scope of the current article, the preliminary
results of this section indicate that further work to disentangle these two
mechanisms would be promising. In the following section, we explore some other
limitations of the current work and provide more recommendations for future
research.

@@

The definition of the domain boundary is only one possible factor that could
influence the effect of contrast on categorization. It is also plausible that
other conditions such as the nature of the instruction (``Generate a new
category that is Not Alpha'' vs ``Generate a new category that is Beta'') and
the boundedness of the features (using features such as hue and orientation that
do have have strict minima or maxima) can affect the extent to which contrast
influences categorization. Though our current contrast models can allow the
influence of contrast to vary, they do not make any predictions regarding the
specific conditions under which contrast can vary. Consequently,
addressing this issue would be an immensely promising avenue for future
research.

>>>>

Creating Something Different: \\
Similarity, Contrast, and Representativeness in Categorization

@@

% ------- ABSTRACT PAGE ------- %
\doublespacing
\section*{Abstract}
The ability to generate new concepts and ideas is among the most fascinating
aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking the extent that the contrast
between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization.

@@

\begin{flushleft}

@@

Creating and learning new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. Computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Yet, the cognitive mechanisms that
enable people to innovate and learn new innovations are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create and learn
new categories that are as interesting as cell phones and smart phones? As a
step towards these goals, in this article we investigate the expectations that
people have when they learn and generate new categories using computational
models and behavioral experiments.

@@

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework that
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another.

@@

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the performance of each model in
capturing human category generation. A third experiment is performed to
highlight the differences between the two contrast models. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

@@

An alternate conceptualization of category contrast is the idea of
representativeness -- exemplars are generated such that they resemble the target
category and are thus more distinct from and less similar to other categories. We
adopt the formalization given by \cite{tenenbaum01}, where the
representativeness $R(y,h)$ of an item $y$ (or in our case, an exemplar) is the
relative amount of evidence that is provided by $y$ for a given hypothesis $h$
in a space of hypotheses $\mathcal{H}$ in contrast to all other hypotheses
$\mathcal{H}^c=\mathcal{H} \setminus \{h\}$:
\begin{equation}
  R(x,h) = \log \dfrac{p(x|h)}{1-p(x|\mathcal{H}^c)} = \log \dfrac{p(x|h)}{\sum_{h'\in H'}{p(x|h')p(h')}} . 
\label{eq:representativeness}
\end{equation}
In Equation \ref{eq:representativeness}, $p(h')$ is the prior probability
distribution on the hypothesis space that excludes $h$. Note that when there are
only two hypotheses (e.g. two categories) involved in the domain, this prior
takes the value of $1$ since there is ever only one alternative hypothesis.

@@

Despite both the original hierarchical Bayesian model and our representativeness
model sharing identical hierarchical structures, their distinct response
processes can yield very different exemplars. Specifically, the hierarchical
Bayesian model emphasizes the generation of exemplars that maintain
distributional commonalities across categories. In contrast, the
representativeness model focuses on the generation of exemplars that are
representative of the underlying distribution for a target category. Generally,
with the assumption of unimodal underlying distributions, this mechanism of
representativeness results in the generation of novel exemplars that are less
similar to exemplars from the known category. This occurs because the
representativeness of novel exemplars tend to be highest where the probability
density of the known category's underlying distribution is the lowest. We see
this illustrated in Figure \ref{fig:representative-examples}, where the
representativeness model displays a strong preference for generating exemplars
from target category 'B' that are further away from the contrast category 'A'.
Exceptions to this pattern can occur (i.e., the generation of novel exemplars
similar to known category exemplars) when the known category distribution is
especially large on all features -- we explore this in more detail in Section
\ref{section:diffcon} where we highlight an important condition demonstrating
how PACKER and the representativeness model can make qualitatively different
predictions.

@@

In the sections below, we present two experiments demonstrating systematic
effects of category contrast on category generation inspired by factors
influencing how the contrast models generate new categories. Our experiments are
based on \cite{jern2013probabilistic}'s paradigm, which is a straightforward
translation of the traditional artificial classification paradigm to the task of
category generation: Participants are first exposed to a single,
experimenter-defined category, and are then asked to generate members of a new
category. We then report quantitative analyses of our results comparing the exemplar dissimilarity
and the representativeness accounts of our results to that of the hierarchical
Bayesian and copy-and-tweak models developed by \cite{jern2013probabilistic}.
After analyzing the category generation studies, we present a third experiment to
distinguish between our two conceptualizations of contrast.

@@

%Contrast model distinction
\section{Experiment 3: Differentiating Contrast}
\label{section:diffcon}
Experiments 1 and 2 clearly established the importance of contrast in category
generation. Follow-up model-based analyses illustrated that both contrast models
account for participant performance better than models that do not take into
account contrast. We also found more support for the hypothesis of contrast as
representativeness versus contrast as exemplar dissimilarity across the
different Alpha conditions.

While the representativeness model fits better than PACKER, and in spite of
their fundamentally different structures, both contrast models seem to make
qualitatively similar predictions. To illustrate this, Figure
\ref{fig:firstbetas} shows how the modal generating probabilities for the first
Beta exemplar are similar across both contrast models.\footnote{Model parameters
  are set to the optimized values from Table \ref{table:global-model-fits}.} The
highest density regions for all models appear to be near the corners and edges.
This is particularly the case for the representativeness model, which almost
exclusively predicts generation at the corners, while PACKER can generate the
first novel exemplars toward the inner regions of the stimulus space in the XOR
condition.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=\textwidth]{figs/firstbetas.pdf}
      \caption{Contrast model predictions indicating probability of generating
        the first Beta exemplar for each Alpha condition. Darker shades of
        blue indicate higher probabilities (normalized within each plot.)}
      \label{fig:firstbetas}
    \end{center}
\end{figure}

These similarities are perhaps unsurprising -- each of the modal generation
regions are maximally distant from the Alpha exemplars. However, recall that
unlike PACKER, the representativeness model does not explicitly use a
distance-based measure in determining exemplar generation probabilities. The
general pattern of producing exemplars dissimilar from the contrast category is
the result of the underlying unimodal category structure assumed by the model
and is not strictly constrained by the distances between exemplars.
Consequently, it is possible to design a configuration of Alpha exemplars where
the representativeness model would predict the generation of novel exemplars
that are \emph{more similar} to the learned exemplars.

In this section, we focus on one particularly useful category type -- the Corner
category -- to distinguish between the two competing accounts of contrast in
category generation. Here, each Alpha exemplar is located at the extreme corners
of the feature space (see Figure \ref{fig:e3-squares}a). We first seek to
identify what predictions each contrast model generates under this condition,
followed by an category generation task, and a brief modeling exercise to
determine which model is a better predictor of the behavioral data.

In the Corner condition, the representativeness model assumes an underlying
Alpha category distribution that is centred on the feature space, with
probability densities highest in the centre and lowest in the corners. Since the
representativeness of novel Beta exemplars generally increase when the
likelihood of the underlying Alpha category distribution decreases, the regions
of highest Beta representativeness are consequently located at the corners where
these Alpha likelihoods are lowest (Figure \ref{fig:e3-squares}b.)

Conversely, PACKER predicts a completely different result. With $\theta_c > 0$,
the probability of generating a Beta exemplar increases with the dissimilarity
from all Alpha exemplars. This results in highest Beta generation probabilities
in the inner regions of the feature space, where the candidate exemplars are
maximally different from the Alpha exemplars (Figure \ref{fig:e3-squares}c.)

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=.8\textwidth]{figs/e3-squares.pdf}
      \caption{Alpha exemplars for the Corner condition (a) and the predictions
        from contrast models PACKER (b) and the representativeness model (c)
        with square stimuli.}
      \label{fig:e3-squares}
    \end{center}
\end{figure}

[Transition to and justification for this still feels a little shaky]

A large part of the representativeness model's predictive accuracy in
Experiments 1 and 2 is due to its preference for generating categories located
at the boundaries of the stimulus features. If at least one of these boundaries
is removed by utilizing an unbounded feature (i.e., no absolute maximum or
minimum exists in the stimulus space along this dimension), we should expect to
see a decrease in the predictive performance of the representativeness model. To
implement this, in our next experiment we also include a new set of stimuli:
circles that vary in size (a bounded feature) and orientation (an
unbounded feature; Figure \ref{fig:e3-circles}a). Before describing the
details of this experiment, we will briefly explore the contrast model
predictions with stimuli that comprise one bounded and one unbounded feature.

\begin{figure}[p]
    \begin{center} 
      \includegraphics[width=.8\textwidth]{figs/e3-circles.pdf}
      \caption{Circle stimulus domain (a) and predictions from contrast models
        PACKER (b) and the representativeness model (c). The unbounded feature
        (orientation) is represented on the y-axis, while the bounded feature
        (size) is represented on a x-axis. As with the square stimuli, feature
        and direction assignment were counterbalanced across participants (in
        the displayed example, circles are rotated clockwise up the y-axis.) }
      \label{fig:e3-circles}
    \end{center}
\end{figure}

Compared to stimuli with two bounded features, PACKER makes a similar prediction
in generating the first Beta exemplar when one feature is unbounded, with the
curious exception that the highest density regions appear compressed along the
y-axis. Due to the unbounded nature of orientation, the maximum distance from
the Alpha exemplars is half of the full range of the feature -- for an Alpha
exemplar oriented at 0 radians, the furthest another exemplar can be situated is
$\pi$ radians within the full range of orientation is $2\pi$ radians.

For the representativeness model, the single unbounded feature essentially
prevents the formation of edge effects along that feature. Consequently, this
results in higher probabilities of the generating the first Beta exemplars along
the edges of the bounded feature (Figure \ref{fig:e3-circles}c). Since this
model does not explicitly recognize similarities as distances along the feature
space (unlike PACKER), it does not experience the compression in density regions
along the unbounded feature axis. Intuitively, this suggests that the
representativeness model (which prefers generating exemplars at the edges) will
be poorer at generating stimuli with an unbounded feature compared to
stimuli with all features bounded.

% We have identified the conditions under which the contrast models may make
% differing predictions, however this distinction is not very useful without any
% empirical support. Data from the next experiment will most certainly support one
% model or the other

[Maybe work on transition if there's time]

We recruited 89 participants from Amazon Mechanical Turk for this experiment, 46
of whom were randomly assigned square stimuli and the remaining 43 assigned
circle stimuli. Square stimuli were exactly the same as in Experiments 1 and 2.
Circle stimuli were made up of unfilled circles with a radial line extending
from its centre to the edge. These stimuli varied along two features -- their
diameter (or size) ranging from 3.0--5.8 cm, and their orientation ranging from
0--$2\pi$ radians. The orientation of each circle stimuli was fixed at a
randomly selected angle for all participants. Similar to Experiments 1 and 2,
participants first observed a training phase where four Alpha exemplars were
presented across three blocks followed by the generation phase. All participants
in this experiment were shown exemplars from the Corner category type.

\subsection{Results and Discussion}

\begin{figure}
    \begin{center}
      \includegraphics[width=\textwidth]{figs/e3-statsboxes.pdf}
    \caption{Box-plots of the distributional statistics from the categories
generated in Experiment 3. Boxes depict the median and quartiles of each
condition, with whiskers placed at 1.5 IQR. All points outside this region are
marked individually. For circle stimuli, the Y Range represents orientation and
the X Range represents size.}
    \label{fig:e3-statsboxes}
    \end{center}
\end{figure}

Overall, we found no significant differences between the distributional statistics of each
generated category between square and circle stimuli (Figure
\ref{fig:e3-statsboxes}. The only exception is seen in the Y range (representing
the orientation of circle stimuli,) which is significantly lower for the circle
stimuli compared to the square stimuli ($t(87) = 4.94$, $p = .001$). This
provides preliminary support for PACKER, which predicts generating exemplars
within a shorter range along the unbounded feature compared to a bounded
feature.

At a rough level (see Table \ref{table:corner_counts}, 35\% of participants who
observed square stimuli tended to generate their first Beta exemplars along the
edges of the feature space. In comparison, only 12\% of participants who
observed circle stimuli generated their first Beta exemplars along the edges of
the feature space. These proportions more closely match the predictions of
PACKER (33\% probability of generating first Beta exemplars at the edges for
squares, and 24\% probability at edges for circles) compared to the
representativeness model (53\% of the former, 49\% probability of the latter).
In addition, these results agree with our prediction that the representativeness
model would be poorer in predicting data from circle stimuli (26\% of
participants observing circle stimuli) compared to square stimuli (39\%
participants observing square stimuli). [No proper stats here... feels a little
dirty. But I suppose there are model fits coming up.]

Quantitative model fits were also obtained by optimizing both contrast models to
the entire dataset including data from Experiments 1 and 2. Fit statistics and
parameter values are presented in Table \ref{table:corner-model-fits}. Overall,
the representativeness model still emerges as the best-fitting model, although
by a much smaller lead compared to fits to only Experiments 1 and
2. At an individual level, 67\% of participants in Experiment 3 were better
fit by PACKER than the representativeness model.

\begin{table} \centering
  \caption{Frequencies of individual participants best fit by each contrast
    model  broken down by stimulus type.}
\label{table:corner_counts}
\begin{tabular}{ l l l l} \\ Stimulus Type & \textbf{PACKER} &
                                                               \textbf{Representativeness}& Total \\
  \hline 
  Squares & 28 & 18 & 46\\
  Circles & 32 & 11 & 43\\
  Total   & 60 & 29 & 89\\
\end{tabular}
\end{table}
\begin{table} \centering
  \caption{Results of model-fitting to the combined datasets from Experiments 1,
    2 and 3. Smaller AIC values correspond to better model fits (adjusted
    for number of parameters.)}
\label{table:corner-model-fits}

\begin{tabular}{ l l} \\ \textbf{PACKER} & \textbf{Representativeness} \\
  \hline $AIC = 12052$ & $AIC = 11991$ \\ 
  $L = -6023$          & $L = -5991$ \\ 
  $c = 0.60$           & $\kappa = 5.91$  \\ 
  $\theta_c = 2.78$    & $\nu = 1.00$ \\
  $\theta_t = 3.17$    & $\lambda = 1.45$ \\
                       &$\theta = 1.86$ \\
\end{tabular}
\end{table}

Ultimately, these results suggest that PACKER's exemplar dissimilarity account
of contrast, as opposed to the representativeness account, is more appropriate
for the Corner category type. On almost every measure of analysis in this
section, PACKER outperformed the representativeness model. Although the
representativeness model best fit the data from Experiments 1 and 2, it
struggled when the learned category occupied the extreme corners of the stimulus
space -- both in cases where the stimuli possess bounded and unbounded features.

[Work on transition time permitting]

@@

We conducted simulations comparing the contrast models' account of our results
to the classical proposals for category generation: a ``copy-and-tweak'' model
(realized as a variant of PACKER with no sensitivity to category contrast), and
a hierarchical Bayesian model designed to explain the classic distributional
similarity effect. In all simulations, we found that the contrast models
captured a previously unexplained and unexplored aspect of human category
generation. In particular, by measuring PACKER's fit as a function of its
prioritization of within- and between-category similarity, we observed that
considering either constraint exclusively results in a relatively low-quality
account. Instead, PACKER's best results were obtained when both constraints are
considered, indicating that human learners do not generate novel concepts
exclusively on the basis of within-category similarity or between
class-contrast. This finding mirrors our behavioral results and demonstrates
that both constraints influence generation when explained with an exemplar
model. Ultimately, our category generation data were best explained by the
representativeness model. The exact nature of its superiority over the other
models is still not yet clear, although we find evidence that it only barely
surpasses PACKER when considering known categories with extreme exemplars.
As we explore in a later section, it is likely that the
representativeness mechanism itself (i.e., independent of the hierarchical
Bayesian foundation) captures a core aspect of both category generation and
category learning.

@@

Although the focus of this article has been to address the role of contrast in category generation, our findings and approach have relevant implications
for research in creative cognition. A central focus of the creative cognition
approach has been to explain acts of creativity in terms of the mental
representations and processes that are commonly studied in cognitive psychology
and cognitive science \citep{finke1992creative,smith1995creative}. However,
unlike other fields in the study of cognition, creative cognition research
rarely employs quantitative models to evaluate the explanatory value of such
representations and processes. Our modeling results provide a concrete example
of how formal approaches may be used to gain insight into the nature of creative
cognition.

@@

\subsection{Exemplar Dissimilarity and Representativeness in Categorization} 
Although both PACKER and the representativeness model perform better than the
classical models, they appear to be capturing fundamentally different aspects of
contrast in categorization. Specifically, while the representativeness model
excelled in predicting our overall category generation data, PACKER was the
best-fitting model in accounting for categories generated from the Corner
category type, where exemplars are located at the corners of the stimulus
domain.

% Given the history of their development, this is perhaps not entirely
% surprising. PACKER is implemented as an extension of the GCM, which was
% originally developed for (and has been highly successful in) category learning
% \citep{nosofsky1986attention, nosofsky1994comparing}. Conversely, the
% representativeness mechanism was conceived as a heuristic describing the
% relative extent to which an individual observation is generated from some
% unobserved population \citep{kahneman1972subjective,abbott16}. This generative
% nature of the representativeness mechanism intuitively lends itself to the
% prediction of category generation data.

@@

Ultimately, the source of the discrepancy between the performance of the two
contrast mechanisms in category generation is still not entirely clear. Our
results could be an indication of a fundamental difference between the processes
involved where exemplar dissimilarity is a larger factor under certain category
configurations while representativeness exerts more influence in others.
Alternatively, it is possible they are both approximating some factor that we
still have not fully identified. While a thorough investigation is beyond the
scope of the current article, the preliminary results of this section indicate
that further work to disentangle these two mechanisms would be promising. In the
following section, we explore some other limitations of the current work and
provide more recommendations for future research.

@@

The definition of the domain boundary is only one possible factor that could
influence the effect of contrast on categorization. It is also plausible that
other conditions such as the nature of the instruction (``Generate a new
category that is Not Alpha'' vs ``Generate a new category that is Beta'') 
can affect the extent to which contrast influences categorization. Though our
current contrast models can allow the influence of contrast to vary, they do not
make any predictions regarding the specific conditions under which contrast can
vary. Consequently, addressing this issue would be an immensely promising avenue
for future research.
